---
title: "Peer grade Project"
author: "Stijn Huysman"
date: "7/1/2018"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
```

1. Loading the data
-------------------

Loading dating from directy
```{r load data}
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
destfile <- "step_data.zip"
download.file(url, destfile)
unzip(destfile)
activity <- read.csv("activity.csv", sep = ",")
```

Giving the name of the variable and assign structure by
```{r activity details}
names(activity)
str(activity)
head(activity[which(!is.na(activity$steps)), ]) # data set with NA rows removed
```
Now the format of the file is ready for analysis.  

2. Histogram of the total number of steps taken each day 
---------------------------------------------------------
3.and Mean and median number of steps taken each day
-----------------------------------------------------------

Group the number of steps by date and intervals. Find the total number of steps per day over all days. Note that some of the days such as 2012-10-01 have no steps data. Remove such rows for this part.

```{r steps mean}
library(reshape2)
activity_melt <- melt(activity[which(!is.na(activity$steps)), ], id.vars = c("date", "interval"))
head(activity_melt)
steps_sum <- dcast(activity_melt, date ~ variable, sum)
head(steps_sum)
```

Then we can find the mean of 'total number of steps per day'.
```{r total steps per day summary}
summary(steps_sum$steps)
```

Histogram of the total number of steps taken each day.
```{r total step per day histogram}
hist(steps_sum$steps, main = "Histogram of total steps taken per day",
     xlab = "Total steps per day", ylab = "Number of days", 
     breaks = 10, col = "steel blue")
abline(v = mean(steps_sum$steps), lty = 1, lwd = 2, col = "red")
abline(v = median(steps_sum$steps), lty = 2, lwd = 2, col = "black")
legend(x = "topright", c("Mean", "Median"), col = c("red", "black"),       lty = c(1, 2), lwd = c(2, 2))
```

Equivalent ggplot.
```{r ggplot histogram}
library(ggplot2)
ggplot(steps_sum, aes(steps)) + geom_histogram(bins = 10)
```

Here's a plot showing the trend in total number of steps taken per day over two months.
```{r total steps trend}

library(lubridate)
steps_sum$date <- as.Date(steps_sum$date)
ggplot(steps_sum, aes(date, steps)) + geom_line() +            
        scale_x_date(date_labels = "%b %d") + 
        ylab("Total number of steps")
```

4. Time series plot of the average number of steps taken and 
---------------------------------------------------------------
5. The 5-minute interval that, on average, contains the maximum number of steps
----------------------------------------------------------------------------

In this section, we make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken averaged across all days.

```{r time series plot}
stepsmeaninterval <- dcast(activity_melt, interval ~ variable, mean, na.rm = TRUE)
head(stepsmeaninterval)
plot(stepsmeaninterval$interval, stepsmeaninterval$steps, ty = "l", 
     xlab = "time interval", ylab = "Average steps", main = "Average 
     steps taken over all days vs \n time interval")
```

The time interval during which the maximum number of steps is taken is

```{r max steps}
maxsteps_interval <- stepsmeaninterval$interval[which.max(stepsmeaninterval$steps)]
maxsteps_interval
```

6. Code to describe and show a strategy for imputing missing data
-----------------------------------------------------------------

First of all, let us get a sense for the missing values. Are there days with all time intervals reporting NA step values? 

We can replace the missing data for a day by the time average over all other days. 
```{r impute missing days data}
## Impute missing values
activity2 <- split(activity, activity$interval)
activity2 <- lapply(activity2, function(x) {
        x$steps[which(is.na(x$steps))] <- mean(x$steps, na.rm = TRUE)
        return(x)
})
activity2 <- do.call("rbind", activity2)
row.names(activity2) <- NULL

activity2 <- split(activity2, activity2$date)
df <- lapply(activity2, function(x) {
        x$steps[which(is.na(x$steps))] <- mean(x$steps, na.rm = TRUE)
        return(x)
})
activity2 <- do.call("rbind", activity2)
row.names(activity2) <- NULL
head(activity2)
```
Assuming that the time intervals form a disjoint partitioning of 24 hrs, i.e. 1 day is found to be erroneous. The time interval for each day corresponds to approximately 40 hours, which refutes the intervals being disjoint.

```{r steps mean2}
library(reshape2)
activity_melt2 <- melt(activity2, id.vars = c("date", "interval"))
steps_sum <- dcast(activity_melt2, date ~ variable, sum, na.rm = TRUE)
head(steps_sum)
```

7. Histogram of the total number of steps taken each day after missing values are imputed
--------------------------------------------------------------------------------------
```{r total step per day histogram2}
hist(steps_sum$steps, main = "Histogram of total steps taken per day", 
     xlab = "Total steps per day", ylab = "Number of days", 
     breaks = 10, col = "steel blue")
abline(v = mean(steps_sum$steps), lty = 1, lwd = 2, col = "red")
abline(v = median(steps_sum$steps), lty = 2, lwd = 2, col = "black")
legend(x = "topright", c("Mean", "Median"), col = c("red", "black"), lty = c(2, 1), lwd = c(2, 2))
```

Number of rows with NA values
```{r NA row num}
sum(is.na(activity$steps))
sum(is.na(activity$steps))*100/nrow(activity) # Percentage of rows with missing values
```

8. Differences in activity patterns: Weekdays vs Weekends
---------------------------------------------------------

Create a new column describing if the date is a weekday or weekend.

```{r weekday status}
library(lubridate)
weekends <- which(weekdays(as.Date(activity2$date)) == "Saturday" |
              weekdays(as.Date(activity2$date)) == "Sunday")
weekdays <- which(weekdays(as.Date(activity2$date)) != "Saturday" &
              weekdays(as.Date(activity2$date)) != "Sunday")
temp <- c(rep("a", length(activity2)))
temp[weekends] <- "weekend"
temp[weekdays] <- "weekday"
length(temp)
names(temp) <- "day"
activity2 <- cbind(activity2, temp)
names(activity2)[4] <- "day"
```

8b. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends
---------------------------------------------------------------------------------------------------
```{r}
activity2split <- split(activity2, activity2$day)
stepsmean_interval <- lapply(activity2split, function(x) {
        temp <- aggregate(x$steps, list(x$interval), mean)
        names(temp) <- c("interval", "steps")
        return(temp)
})

## Unsplit stepsmean_interval
stepsmean_interval <- do.call("rbind", stepsmean_interval)
weekdays <- grep("weekday" ,row.names(stepsmean_interval))
weekends <- grep("weekend" ,row.names(stepsmean_interval))
temp <- c(rep("a", length(stepsmean_interval$steps)))
temp[weekdays] <- "weekdays"
temp[weekends] <- "weekends"
stepsmean_interval <- cbind(stepsmean_interval, temp)
row.names(stepsmean_interval) <- NULL
names(stepsmean_interval)[3] <- "day"
head(stepsmean_interval)
tail(stepsmean_interval)

library(ggplot2)
ggplot(stepsmean_interval, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) 
```

The mean number of steps taken over the weekdays and weekends.

```{r weekday/end mean steps}
stepsdatamelt <- melt(stepsmean_interval, id.vars = c("interval",
                                                      "day"))
dcast(stepsdatamelt, day ~ variable, mean) # Average steps
```






